Statistical Machine Learning Models on A Bank Marketing dataset

### 2025-12-08 Slides.Rmd
eda slides
Update slides

---
title: "Statistical Machine Learning Models on A Bank Marketing dataset"
author: "Sandra, Tigist, Mussa Hassen, Waddah"
subtitle: 'Colorado State University, DSCI 445: Statistical Machine Learning, Dr.Andee
  Kaplan'
output:
  ioslides_presentation: 
    css: styles.css

@@

@@

## Project Motivation & Business Problem
* **Discuss:** The bank's need for efficient marketing and the high cost of telemarketing campaigns. The goal is to maximize the success rate by predicting customer subscription.
* **Discuss:** The challenge of less positive class (imbalanced data) and why prediction accuracy is crucial.
* **Image/Plot Idea:** A simple, high-level info graphic of a marketing campaign funnel, highlighting the bottleneck/drop-off point.

[Image of a marketing funnel illustrating customer conversion]

@@

## Methodology: Data Preparation (Slide 5)
* **Discuss:** **Data Wrangling** involved converting categorical variables to factors and strategies for handling 'unknown' levels (treating them as a separate category vs. missing).
* **Image/Plot Idea:** A simple flow chart showing: Raw Data $\rightarrow$ Wrangling/Cleaning $\rightarrow$ Train/Test Split. 
* **Image/Plot Idea:** Summary stats or informational plots

[Image of a data science workflow diagram]

@@

## Boosting 1/3: Model Formulation
* **Discuss:** **Boosting** (e.g., Gradient Boosting) is an ensemble method that sequentially builds weak learners (trees) to minimize the classification error of the previous step.
* **Discuss:** This model is expected to be our **highest-performing benchmark** but sacrifices interpretability due to its complex, additive structure.
* **Image/Plot Idea:** A conceptual diagram illustrating the boosting process (weak learners combining sequentially to form a strong learner). 
```{r}
library(gbm)

bank_train$y_num <- ifelse(bank_train$y == "yes", 1, 0)
bank_test$y_num  <- ifelse(bank_test$y  == "yes", 1, 0)

boost_model <- gbm(
  formula = y_num ~ . - y,       
  data = bank_train,
  distribution = "bernoulli",   
  n.trees = 1000,                
  interaction.depth = 3,         
  shrinkage = 0.01,             
  n.minobsinnode = 10,
  verbose = FALSE
)

boost_prob <- predict(
  boost_model,
  newdata = bank_test,
  n.trees = 1000,
  type = "response"
)

boost_pred <- ifelse(boost_prob > 0.5, "yes", "no")

boost_accuracy <- mean(boost_pred == bank_test$y)
boost_error    <- 1 - boost_accuracy

cat("Boosting Test Accuracy:", round(boost_accuracy * 100, 2), "%\n")
cat("Boosting Test Error:",   round(boost_error * 100, 2), "%\n")

plot(
  boost_model$train.error,
  type = "l",
  xlab = "Number of Trees (Iterations)",
  ylab = "Training Loss",
  main = "Boosting: Sequential Error Reduction"
)

## Boosting 2/3: Hyperparameter Tuning 
* **Discuss:** Multiple hyperparameters were tuned using **Cross-Validation**, including the **number of trees** (M), **interaction depth** (complexity of weak learners), and **learning rate** ($\eta$).
* **Discuss:** The tuning process involves finding the right balance between these parameters to ensure the model generalizes well without overfitting the training data.
* **Image/Plot Idea:** A plot showing the **Error Rate vs. Number of Trees**, highlighting where the Test Error flattens or begins to increase (overfitting).
```{r}
boost_model <- gbm(
  formula = y_num ~ . - y,
  data = bank_train,
  distribution = "bernoulli",
  n.trees = 1000,
  interaction.depth = 3,
  shrinkage = 0.01,
  n.minobsinnode = 10,
  verbose = FALSE
)

M_values <- seq(50, 1000, by = 50)
test_error <- numeric(length(M_values))

for (i in seq_along(M_values)) {
  M <- M_values[i]

  prob_M <- predict(
    boost_model,
    newdata = bank_test,
    n.trees = M,
    type = "response"
  )

  pred_M <- ifelse(prob_M > 0.5, "yes", "no")
  test_error[i] <- mean(pred_M != bank_test$y)
}

best_idx <- which.min(test_error)
best_M   <- M_values[best_idx]

cat("Best number of trees based on Test Error:", best_M, "\n")

plot(
  M_values, test_error,
  type = "b",
  xlab = "Number of Trees (M)",
  ylab = "Test Error",
  main = "Boosting: Test Error vs Number of Trees"
)

abline(v = best_M, lty = 2)

## Boosting 3/3: Evaluation & Importance 
* **Discuss:** The final Boosting model was evaluated on the **Test Set**. Confirm that it achieved the **highest AUC**, setting the benchmark for predictive accuracy.
* **Discuss:** While coefficients are not available, discuss the model's **Variable Importance Plot** as the only glimpse into its internal workings.
* **Image/Plot Idea:** The **Variable Importance Plot** from the Boosting model, which should confirm that the most important features align reasonably well with LogReg's significant coefficients.
```{r}
library(pROC)

boost_prob <- predict(
  boost_model,
  newdata = bank_test,
  n.trees = 1000,       
  type = "response"
)

test_y_fac <- factor(bank_test$y, levels = c("no", "yes"))

boost_roc <- roc(
  response  = test_y_fac,
  predictor = boost_prob,
  levels    = c("no", "yes")
)

boost_auc <- as.numeric(auc(boost_roc))
cat("Boosting Test AUC:", round(boost_auc, 3), "\n")

vi <- summary(boost_model, plotit = FALSE)

top_n <- min(10, nrow(vi))
vi_top <- vi[1:top_n, ]

barplot(
  height = vi_top$rel.inf,
  names.arg = vi_top$var,
  las = 2,
  ylab = "Relative Influence",
  main = "Boosting: Variable Importance (Top Features)"
)

## Results: Comparative Evaluation (Slide 13). #MUSSA
* **Discuss:** Present the **Final Summary Table** comparing the Test Set performance (AUC, Recall, Accuracy) of all models (**Best LogReg**, **KNN**, **Boosting**).
* **Discuss:** Visually compare the models using the **Combined ROC Curve Plot** and discuss which model dominates the space.
* **Image/Plot Idea:** The **Combined ROC Curve Plot** comparing all three final models (LogReg, KNN, and Boosting).

>>>>

---
title: "Statistical Machine Learning Models on A Bank Marketing dataset"
author: "Sandra, Tigist, Mussa Hassen, Waddah"
subtitle: 'Colorado State University, DSCI 445: Statistical Machine Learning, Dr.Andee
  Kaplan'
output:
  ioslides_presentation:
    css: styles.css
---

@@

## boosting

if ("residuals" %in% colnames(train_data)) {
  train_data$residuals <- NULL
}

train_data$y_num <- ifelse(train_data$y == "yes", 1, 0)
test_data$y_num  <- ifelse(test_data$y == "yes", 1, 0)

library(gbm)
set.seed(445)

boost_model <- gbm(
  y_num ~ . - y,         
  data = train_data,
  distribution = "bernoulli",
  n.trees = 1500,       
  shrinkage = 0.01,
  interaction.depth = 3,
  bag.fraction = 0.5,
  verbose = FALSE
)
best_trees <- gbm.perf(boost_model, method = "OOB", plot.it = FALSE)

boost_prob <- predict(boost_model, test_data, n.trees = best_trees, type = "response")
boost_pred <- factor(ifelse(boost_prob > 0.5, "yes", "no"), levels = c("no","yes"))

cm_boost <- confusionMatrix(boost_pred, test_y, positive = "yes")
cm_boost

roc_boost <- roc(test_y, boost_prob)
auc_boost <- auc(roc_boost)

## test 
final_results <- data.frame(
  Model = c("Logistic", "Ridge", "LASSO", "KNN", "GAM", "Boosting"),
  Accuracy = c(
    cm_baseline$overall["Accuracy"],
    cm_ridge$overall["Accuracy"],
    cm_lasso$overall["Accuracy"],
    cm_knn$overall["Accuracy"],
    cm_gam$overall["Accuracy"],
    cm_boost$overall["Accuracy"]
  ),
  Sensitivity = c(
    cm_baseline$byClass["Sensitivity"],
    cm_ridge$byClass["Sensitivity"],
    cm_lasso$byClass["Sensitivity"],
    cm_knn$byClass["Sensitivity"],
    cm_gam$byClass["Sensitivity"],
    cm_boost$byClass["Sensitivity"]
  ),
  AUC = c(
    auc(roc_baseline),
    auc(roc_ridge),
    auc(roc_lasso),
    auc_knn,
    auc_gam,
    auc_boost
  )
)
# combined ROC 
roc_all <- rbind(
  data.frame(fpr = 1 - roc_baseline$specificities, tpr = roc_baseline$sensitivities, Model = "Logistic"),
  data.frame(fpr = 1 - roc_ridge$specificities,    tpr = roc_ridge$sensitivities,    Model = "Ridge"),
  data.frame(fpr = 1 - roc_lasso$specificities,    tpr = roc_lasso$sensitivities,    Model = "LASSO"),
  data.frame(fpr = 1 - roc_knn$specificities,      tpr = roc_knn$sensitivities,      Model = "KNN"),
  data.frame(fpr = 1 - roc_gam$specificities,      tpr = roc_gam$sensitivities,      Model = "GAM"),
  data.frame(fpr = 1 - roc_boost$specificities,    tpr = roc_boost$sensitivities,    Model = "Boosting")
)

@@

## Project Motivation & Business Problem (S)
- Telemarketing campaigns are expensive, time-consuming, and have very low success rates. The bank wants to improve efficiency by identifying which clients are most likely to subscribe.
- The dataset has a severe class imbalance: only a small fraction of customers subscribe to the term deposit. This makes prediction difficult and increases the risk of false negatives.
- Improving predictive accuracy would allow the bank to target fewer customers, reduce costs, and increase return on investment (ROI) for future campaigns.

```{r, echo=FALSE}
library(ggplot2)

ggplot(bank_full, aes(x = y, fill = y)) +
geom_bar() +
scale_fill_manual(values = c("#2E86C1", "#F39C12")) +
labs(
title = "Term Deposit Subscription (Severe Class Imbalance)",
x = "Subscribed?",
y = "Count"
) +
theme_minimal(base_size = 14)

@@

## Methodology: Data Preparation (Slide 5) (s)

- Converted all categorical variables into factors so models interpret them correctly.
- Addressed "unknown" values by treating them as their own category to avoid losing information.
- Created a new feature previously_contacted to properly encode pdays.
- Split the dataset into 80% training and 20% test using stratified sampling to preserve the class imbalance.
- Ensured numeric predictors were available for later scaling, tuning, and model evaluation.

## Breakdown of Categorical Variable
```{r, echo=FALSE}
library(ggplot2)

ggplot(bank_full, aes(x = job)) +
geom_bar(fill = "steelblue") +
coord_flip() +
labs(
title = "Distribution of Job Categories",
x = "Job",
y = "Count"
) +
theme_minimal()

@@

## Boosting 1/3: Model Formulation
- Gradient Boosting Machines (GBM) build an ensemble of weak learners (typically shallow decision trees).
- Each tree is fit to the errors (residuals) of the previous tree — allowing the model to improve sequentially.
- This iterative correction makes boosting highly flexible and able to capture complex, nonlinear patterns in the data.
- Because the model is effectively a sum of many small trees, it becomes a strong predictor but loses interpretability relative to logistic regression.
- Boosting is expected to be our highest-performing model, especially on imbalanced datasets like this one.

## Boosting 2/3: Hyperparameter Tuning 
- We tuned the number of trees using test-error curves to prevent overfitting.
- Test error decreased steadily and leveled off around M = 1500 trees, which we selected as the optimal value.
- This choice balances model complexity and generalization, giving strong predictive performance without excessive tree growth.

## Boosting 3/3: Evaluation & Importance

## Results: Comparative Evaluation (Slide 13). (S)
- We evaluated six models using Accuracy, Sensitivity (Recall), and AUC on the Test Set.
- Boosting achieved the highest overall accuracy (0.905) and the highest sensitivity (0.412), meaning it identified the most true positives.
- Logistic, Ridge, and LASSO models performed very similarly in accuracy (~0.898–0.900) and AUC (~0.907–0.908), but with lower sensitivity.
GAM achieved moderate accuracy and AUC but very low sensitivity.
KNN performed the worst, with the lowest AUC (0.617), indicating weak separation between classes.

```{r, echo=FALSE}
final_results <- data.frame(
  Model = c("Logistic", "Ridge", "LASSO", "KNN", "GAM", "Boosting"),
  Accuracy = c(
    cm_baseline$overall["Accuracy"],
    cm_ridge$overall["Accuracy"],
    cm_lasso$overall["Accuracy"],
    cm_knn$overall["Accuracy"],
    cm_gam$overall["Accuracy"],
    cm_boost$overall["Accuracy"]
  ),
  Sensitivity = c(
    cm_baseline$byClass["Sensitivity"],
    cm_ridge$byClass["Sensitivity"],
    cm_lasso$byClass["Sensitivity"],
    cm_knn$byClass["Sensitivity"],
    cm_gam$byClass["Sensitivity"],
    cm_boost$byClass["Sensitivity"]
  ),
  AUC = c(
    auc(roc_baseline),
    auc(roc_ridge),
    auc(roc_lasso),
    auc_knn,
    auc_gam,
    auc_boost
  )
)

knitr::kable(final_results, digits = 3)

```
---

```{r, echo=FALSE}
roc_all <- rbind(
  data.frame(fpr = 1 - roc_baseline$specificities, tpr = roc_baseline$sensitivities, Model = "Logistic"),
  data.frame(fpr = 1 - roc_ridge$specificities,    tpr = roc_ridge$sensitivities,    Model = "Ridge"),
  data.frame(fpr = 1 - roc_lasso$specificities,    tpr = roc_lasso$sensitivities,    Model = "LASSO"),
  data.frame(fpr = 1 - roc_knn$specificities,      tpr = roc_knn$sensitivities,      Model = "KNN"),
  data.frame(fpr = 1 - roc_gam$specificities,      tpr = roc_gam$sensitivities,      Model = "GAM"),
  data.frame(fpr = 1 - roc_boost$specificities,    tpr = roc_boost$sensitivities,    Model = "Boosting")
)

ggplot(roc_all, aes(x = fpr, y = tpr, color = Model)) +
  geom_line(size = 1) +
  geom_abline(linetype = "dashed") +
  labs(
    title = "ROC Curve Comparison for All Models",
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  theme_minimal()
```
